Analyzing Student Use of Spacing and Interleaving Strategies in Interactions with GenAI-Powered Chatbots in Programming Courses

### 2024-11-18 CSEDU2025/Paper.tex
update

In response to the growing need for deeper insight into the use of GenAI tools
in educational settings, this study aims to help address this gap by analyzing
the interactions between students in a Java programming course and an
educational chatbot powered by OpenAI gpt-3.5, enhanced with the
Retrieval-Augmented Generation (RAG) technique. Specifically, it focuses on
examining these interactions within the context of two metacognitive strategies:
spacing and interleaving. To achieve this objective, we formulate three research
questions:

@@

This section outlines the theoretical framework that grounds the study. We
begin by introducing the concept of metacognition, along with the spacing and
interweaving strategies applied in this research. Next, we examine related works
 organized into two subsections: the use of GenAI in programming and recent
 studies emphasizing the importance of metacognition in the use of GenAI.

@@

\subsection{GenAI and Programming}

@@

\subsection{GenAI and Metacognition}

\cite{Sidra24} addressed the growing importance of metacognitive skills in an
AI-enabled workforce. The problem discussed was how AI systems, especially
collaborative tools like Large Language Models (LLMs), impacted human
decision-making and the challenges they posed due to the lack of human-like
metacognition in AI. The method used identified four main characteristics that
differentiated human-AI interaction from human-human collaboration and proposed,
based on the dual-process theory of human thought, that metacognitive thinking
was crucial for the effective use of AI. The results indicated that human
workers needed to develop and apply metacognitive skills, such as planning,
monitoring, and evaluating their thought processes, to mitigate cognitive biases
and improve decision-making when working with AI. The study concluded by
recommending improvements in AI design and training to improve workers'
metacognitive abilities, ensuring more effective collaboration with AI systems.

Learning based on metacognitive strategies has been shown to be effective in
enhancing students’ performance and skills. \cite{Zheng19} investigated the
effects of metacognitive scaffolds on group behavior, performance, and cognitive
load in computer-supported collaborative environments. The results showed
positive impacts on metacognitive behavior and group performance without
increasing the cognitive load. \cite{LiWei23} proposed a metacognition-based
collaborative approach to enhance performance in collaborative programming. The
findings indicated that this approach influenced computational thinking,
critical thinking, and metacognitive awareness. Similarly, \cite{Wang23}
explored how metacognitive instruction affects computational thinking, critical
thinking, and metacognitive skills in collaborative programming, observing
improvements in problem-solving and collaboration. \cite{Khusnul24} investigates
the efficacy of a meta-learning approach in improving metacognitive and creative
skills. This study confirms the effectiveness of metacognition strategies and
elucidates the relationship between meta-learning and metacognition. This
research suggests that metacognition can improve metacognitive abilities,
providing valuable insights into educational technology and course design in
higher education settings.

>>>>

In response to the growing need for deeper insight into the use of GenAI tools
in educational settings, this study aims to help address this gap by analyzing
the interactions between students in a Java programming course and an
educational chatbot powered by OpenAI gpt-3.5, enhanced with the
Retrieval-Augmented Generation (RAG) technique. Specifically, it focuses on
examining these interactions within the context of two metacognitive strategies:
spacing \citep{Carvalho20} and interleaving \citep{Rivers21}. To achieve this
objective, we formulate three research questions:

@@

This section outlines the theoretical framework that is the basis of the study.
We begin by introducing the concept of metacognition, along with the spacing and
interweaving strategies applied in this research. Next, we present related works
organized into two subsections: research involving GenAI and studies focusing
on metacognition.

@@

\subsection{GenAI Related Works}

@@

\subsection{Metacognition Related Works}

Learning based on metacognitive strategies has been shown to be effective in
enhancing students’ performance and skills. \cite{Zheng19} investigated the
effects of metacognitive scaffolds on group behavior, performance, and cognitive
load in computer-supported collaborative environments. The results showed
positive impacts on metacognitive behavior and group performance without
increasing the cognitive load. \cite{LiWei23} proposed a metacognition-based
collaborative approach to enhance performance in collaborative programming. The
findings indicated that this approach influenced computational thinking,
critical thinking, and metacognitive awareness. Similarly, \cite{Wang23}
explored how metacognitive instruction affects computational thinking, critical
thinking, and metacognitive skills in collaborative programming, observing
improvements in problem-solving and collaboration. \cite{Khusnul24} investigates
the efficacy of a meta-learning approach in improving metacognitive and creative
skills. This study confirms the effectiveness of metacognition strategies and
elucidates the relationship between meta-learning and metacognition. This
research suggests that meta-learning can improve metacognitive abilities,
providing valuable insights into educational technology and course design in
higher education settings.

\cite{Sidra24} addressed the growing importance of metacognitive skills in an
AI-enabled workforce. The problem discussed was how AI systems, especially in
collaborative tools based on Large Language Models (LLMs), impacted human
decision-making and the challenges they posed due to the lack of human-like
metacognition in AI. The method used identified four main characteristics that
differentiated human-AI interaction from human-human collaboration and proposed,
based on the dual-process theory of human thought, that metacognitive thinking
was crucial for the effective use of AI. The results indicated that human
workers needed to develop and apply metacognitive skills, such as planning,
monitoring, and evaluating their thought processes, to mitigate cognitive biases
and improve decision-making when working with AI. The study concluded by
recommending improvements in AI design and training to improve workers'
metacognitive abilities, ensuring more effective collaboration with AI systems.
